Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-17 E-Hypergraphs-LICS24.tex
minor fixes

CHRIS: FIX BELOW DEFINITION. 
\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item all elements in the image of $i+j$ share the same set of parents;
        \item $m(v)  \consistency m(w)$ for all $v,w$ in the image of $i+j$.
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces are given by the following:
		\begin{align*}
		&n'' = (i'' \setminus i) + l\\
		&k'' = (j'' \setminus j) + r\\
		&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
		&f_2 : l \to \mathcal{C} \to \mathcal{H} 
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
Since both $f_1$ and $f_2$ are monos, their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The conditions (3) and (4) on a match, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]

With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense.

@@

We wish to specify a set of EDPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of $\Sigma$-terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket f ; g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} ; \llbracket g \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}

>>>>

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces are given by the following:
		\begin{align*}
		&n'' = (i'' \setminus i) + l\\
		&k'' = (j'' \setminus j) + r\\
		&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
		&f_2 : l \to \mathcal{C} \to \mathcal{H} 
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
Since both $f_1$ and $f_2$ are monos with disjoint image (?), their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.
CHRIS: FIX (4) ABOVE:

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
%The conditions above, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
%\[
%	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
%\]
%We use 
With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense.

@@

We wish to specify a set of EDPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of $\Sigma$-terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket f ; g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} ; \llbracket g \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  It uses the notation of dashed blue boxes for hierarchical edges, with its children drawn inside,  and maximally consistent components seperated by another dashed line.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}
